---
title: Linting
---

Panache includes a built-in linter that checks for correctness issues and best
practice violations in your Markdown documents. Unlike the formatter which
handles style consistency, the linter catches semantic problems like syntax
errors, heading hierarchy issues, broken references, and citation problems.

## Philosophy

The linter focuses on semantic correctness rather than stylistic preferences,
for which the formatter is responsible (and can act as a linter via
`panache format --check`).

## CLI Usage

### Basic Linting

To lint a single file and show diagnostics, run this:

```bash
panache lint document.qmd
```

To lint multiple files at once, you can specify them one-by-one:

```bash
panache lint file1.md file2.qmd file3.Rmd
```

If you want to lint all supported files in a directory, use:

```bash
panache lint .
panache lint docs/
```

`panache lint` supports glob patterns as well:

```bash
panache lint 'src/**/*.qmd'
```

### Lint from stdin

You can also pipe content through the linter:

```bash
cat document.qmd | panache lint # or `panache lint < document.qmd`
```

Or with a here-document:

```bash
panache lint <<'EOF'
# H1
### H3
EOF
```

### Apply Automatic Fixes

Panache can automatically fix certain types of issues:

```bash
panache lint --fix document.qmd
```

::: {.callout-warning}
Auto-fix modifies files in place. Always commit your changes first or use
version control.
:::

### CI Mode

If you want to enforce linting in a CI/CD pipeline, use the `--check` flag. This
will run the linter and exit with code 1 if any violations are found:

```bash
panache lint --check .
```

This mode is ideal for enforcing linting in CI/CD pipelines:

```bash
panache lint --check . || exit 1
```

### Message format

Use `--message-format` to control diagnostic verbosity:

```bash
panache lint --message-format human document.qmd # default rich output
panache lint --message-format short document.qmd # compact one-line diagnostics
```

## Lint Rules

Panache includes several built-in lint rules that analyze document structure and
content.

### `heading-hierarchy`

Detects skipped heading levels that violate document structure best practices.

Severity
:   Warning

Auto-fix
:   Yes

Description
:   Headings should increment by at most one level (e.g., H1 → H2 → H3).
    Skipping levels (H1 → H3) makes the document structure unclear and can break
    table-of-contents generation.

**Example violation:**

```markdown
# Main Title

### Subsection
```

**Diagnostic:**

```
warning: [heading-hierarchy] Heading level skipped from h1 to h3; expected h2
 --> document.qmd:3:1
  |
3 | ### Subsection
  | ^^^^^^^^^^^^^^
```

**Auto-fix:** Changes `### Subsection` to `## Subsection`.

**Valid structure:**

```markdown
# Main Title

## Section

### Subsection
```

### `duplicate-reference-labels`

Detects duplicate reference link and footnote definitions.

Severity
:   Warning

Auto-fix
:   No

Description
:   Each reference label and footnote ID must be unique within a document.
    Duplicate definitions cause ambiguity—only the first definition is used,
    making the others ineffective.

**Example violation**

```markdown
See [link1] and [link2].

[link1]: https://example.com
[link1]: https://different.com
```

**Diagnostic:**

```
warning[duplicate-reference-labels]: Duplicate reference definition 'link1'
  --> document.qmd:4:1
note: First defined here:
  --> document.qmd:3:1
```

**Footnote example:**

```markdown
Text with footnote[^1] and another[^2].

[^1]: First footnote.
[^1]: Duplicate footnote.
```

**Resolution:** Rename or remove the duplicate definition.

### `undefined-references`

Detects reference links and footnotes that point to missing definitions.

Severity
:   Warning

Auto-fix
:   No

Description
:   Flags unresolved reference-style links (including shortcut/collapsed forms)
    and unresolved footnote references. This helps catch broken cross-references
    early in editing and CI.

**Example violation:**

```markdown
See [missing][nope] and note[^missing].

[ok]: https://example.com
```

**Diagnostic:**

```
warning[undefined-reference-label]: Reference label '[nope]' not found
  --> document.qmd:1:15
warning[undefined-footnote-id]: Footnote '[^missing]' not found
  --> document.qmd:1:31
```

### `citation-keys`

Validates citation keys against loaded bibliographies and detects conflicts in
inline bibliography entries.

Severity
:   Error for loading failures, Warning for undefined keys

Auto-fix
:   No

Requirements
:   Requires `extensions.citations = true` in configuration

Description
:   Checks that all cited keys (`[@key]`) exist in the configured bibliography
    files. Also validates inline bibliography entries for duplicates and
    conflicts.

**Example violation (undefined key):**

```markdown
---
bibliography: refs.bib
---

See @smith2020 and @jones2021.
```

If `jones2021` doesn't exist in `refs.bib`:

```
warning[citation-keys]: Citation key 'jones2021' not found in bibliography
  --> document.qmd:6:17
```

**Example violation (bibliography load error):**

```markdown
---
bibliography: nonexistent.bib
---
```

**Diagnostic:**

```
error[bibliography-load-error]: Failed to load bibliography nonexistent.bib: File not found
  --> document.qmd:1:1
```

**When it runs:** Only when document metadata includes bibliography
configuration and citation extension is enabled.

### `chunk-label-spaces`

Detects executable chunk labels containing whitespace (for example
`{r several words}` or `label="several words"`).

Severity
:   Warning

Auto-fix
:   No

Description
:   Labels with spaces are accepted by Quarto execution, but cross-references
    often fail to resolve reliably. Use a stable identifier such as
    `several-words` or `several_words` instead.

### `missing-chunk-labels`

Detects executable chunks that do not define a `label` (either inline or
hashpipe style).

Severity
:   Warning

Auto-fix
:   No

Description
:   Labels facilitates debugging. Add a label with either `#| label: my-chunk`
    or inline `label=my-chunk`.

### `figure-crossref-captions`

Detects figure cross-references that point to chunk labels without a figure
caption option.

Severity
:   Warning

Auto-fix
:   No

Description
:   Bookdown figure cross-references (`\@ref(fig:...)`) require a captioned
    chunk to create a resolvable figure label at render time. When the target
    chunk has a `label` but no `fig-cap`/`fig.cap`, the crossref will not
    resolve.

### `unknown-emoji-alias`

Detects `:alias:` emoji shortcodes that are not recognized.

Severity
:   Warning

Auto-fix
:   No

Requirements
:   Requires `extensions.emoji = true` in configuration

Description
:   Checks parsed emoji aliases against the emoji shortcode dataset and warns
    when an alias is unknown.

**Example violation:**

```markdown
Looks good :smile:, but this one is wrong :not-a-real-emoji:.
```

**Diagnostic:**

```
warning[unknown-emoji-alias]: Unknown emoji alias ':not-a-real-emoji:'
  --> document.qmd:1:40
```

## YAML parse diagnostics

Panache emits `yaml-parse-error` diagnostics when embedded YAML content is
invalid. This currently applies to both document frontmatter (`--- ... ---`) and
executable chunk hashpipe options (`#| ...`).

Severity
:   Warning

Auto-fix
:   No

**Example (hashpipe):**

````markdown
```{{r}}
#| echo: [
1 + 1
```
````

**Diagnostic:**

```
warning[yaml-parse-error]: YAML parse error: ...
  --> document.qmd:2:10
```

## Diagnostic Format

Diagnostics are displayed in a compiler-style format:

```
severity[rule-name]: message
  --> file:line:column
```

Components:

`severity`
:   `error`, `warning`, or `info`

`rule-name`
:   The lint rule that triggered (e.g., `heading-hierarchy`)

`message`
:   Human-readable description of the issue

`location`
:   File path, line number, and column number

**Multi-location diagnostics:**

Some rules provide additional context:

```
warning[duplicate-reference-labels]: Duplicate reference definition 'link1'
  --> document.qmd:4:1
note: First defined here:
  --> document.qmd:3:1
```

## Severity Levels

Panache uses three severity levels:

`error`
:   Critical issues that prevent correct parsing or rendering

`warning`
:   Likely mistakes or best practice violations

`info`
:   Informational messages (currently unused, reserved for future use)

## Auto-Fix Capabilities

Auto-fixes are available for select rules where the correction is unambiguous:

Currently supported:

- `heading-hierarchy`: Adjusts heading levels to fix skipped levels

Currently unsupported:

- `duplicate-reference-labels`: Multiple valid resolutions (rename, delete,
  merge)
- `citation-keys`: Requires external bibliography editing
- Parser errors: Too context-dependent

::: {.callout-note}
Auto-fixes are applied in place when using `--fix`. Always use version control
or backup files before applying automatic fixes to important documents.
:::

## External Linters

Panache can integrate with external code linters to check code blocks within
your documents.

### Configuration

Enable external linters in your configuration file:

```toml
[linters]
r = "jarl"
python = "ruff"
sh = "shellcheck"
js = "eslint"
go = "staticcheck"
rust = "clippy"
```

Available external linters:

| Language              | Linter        | Description                         |
| --------------------- | ------------- | ----------------------------------- |
| R                     | `jarl`        | R linter with JSON diagnostics      |
| Python                | `ruff`        | Python linter with JSON diagnostics |
| Shell                 | `shellcheck`  | Shell linter with JSON diagnostics  |
| JavaScript/TypeScript | `eslint`      | JS/TS linter with JSON diagnostics  |
| Go                    | `staticcheck` | Go linter with JSON diagnostics     |
| Rust                  | `clippy`      | Rust linter with JSON diagnostics   |

### How External Linters Work

External linters analyze code blocks using a stateful concatenation approach:

1. **Collection**: All code blocks of the target language are extracted
2. **Concatenation**: Blocks are joined with blank-line preservation to maintain
   line numbers
3. **Analysis**: The external linter analyzes the concatenated code
4. **Mapping**: Diagnostics are mapped back to original document positions

::: {.callout-note}
This approach correctly handles stateful code. If a variable is defined in one
code block and used in another, the linter sees the complete context and won't
report false positives.
:::

### Example

Document with multiple R code blocks:

````markdown
---
title: "Analysis"
---

```{r}
x <- 10
```

Some text between blocks.

```{r}
y <- x + 5
```
````

With `[linters] r = "jarl"` configured, `jarl` analyzes both blocks together and
correctly understands that `x` is defined before it's used. Similarly, with
`[linters] python = "ruff"`, Ruff can lint Python code blocks and map
diagnostics back to the original document.

### Behavior

Language matching
:   Case-insensitive matching of code block language to linter configuration

Error handling
:   Missing linter executables log a warning and skip gracefully

Compatibility checks
:   External linters only run for their supported languages. Unsupported
    mappings in `[linters]` (for example, `bash = "jarl"`) are skipped with a
    warning.

Timeout
:   30-second timeout per linter invocation

Line accuracy
:   Diagnostics report exact line and column positions in the original document

Auto-fixes
:   Supported for external linters that return fix edits with mappable ranges
    (currently `jarl`, `ruff`, and `eslint`)

### Where External Linters Run

CLI
:   Diagnostics appear in `panache lint` output

LSP
:   Diagnostics appear inline in your editor with live updates

## Ignore Directives

You can selectively disable linting for specific regions using HTML comment
directives:

### Ignore Linting Only

Use `panache-ignore-lint-start` and `panache-ignore-lint-end` to suppress lint
warnings:

```markdown
Normal content will be linted.

<!-- panache-ignore-lint-start -->
#### This heading skip won't trigger heading-hierarchy warning
<!-- panache-ignore-lint-end -->

Back to normal linting.
```

This is useful for:

- Intentional heading level skips for specific formatting
- Generated content with unusual structure
- Third-party content you don't control
- Documentation examples showing bad practices

#### Ignore Both Formatting and Linting

Use `panache-ignore-start` and `panache-ignore-end` to disable both:

```markdown
<!-- panache-ignore-start -->
#### Unusual   structure   here
Both formatting and linting ignored in this region
<!-- panache-ignore-end -->
```

::: {.callout-note}
**Note on Directive Behavior**: Lint rules still "see" content in ignored
regions when tracking context (e.g., for heading hierarchy), but diagnostics
from ignored regions are filtered out. This ensures rules maintain proper state
across the document.
:::

See [Formatting](formatting.qmd) for more information about formatting-specific
ignore directives.

## Configuration

Lint rules can be configured in the `[lint.rules]` section of your configuration
file. Each key is a rule name and each value is a boolean (`true`/`false`).

```toml
[lint.rules]
heading-hierarchy = true
duplicate-reference-labels = true
undefined-references = true
citation-keys = true
chunk-label-spaces = true
missing-chunk-labels = true
figure-crossref-captions = true
unknown-emoji-alias = true
```

All rules are enabled by default when omitted. You can disable a specific rule:

```toml
[lint.rules]
undefined-references = false
```

::: {.callout-note}
Legacy `[lint] rule = true/false` is still supported for backward compatibility,
but deprecated.
:::

## LSP Integration

When using the Panache language server, lint diagnostics appear live in your
editor as you type:

- Squiggly underlines for errors and warnings
- Hover tooltips showing diagnostic messages
- Code actions for auto-fixes (e.g., fix heading hierarchy)

See the [LSP documentation](lsp.qmd) for editor configuration details.

## Examples

### Example 1: Heading Hierarchy

Before:

```markdown
# Main Title

### Skipped Level

#### Another Skip
```

Run linter:

```bash
panache lint document.qmd
```

Output:

```
warning: [heading-hierarchy] Heading level skipped from h1 to h3; expected h2
 --> document.qmd:3:1
  |
3 | ### Skipped Level
  | ^^^^^^^^^^^^^^^^^^

warning: [heading-hierarchy] Heading level skipped from h3 to h4; expected h3
 --> document.qmd:5:1
  |
5 | #### Another Skip
  | ^^^^^^^^^^^^^^^^^
```

Apply auto-fix:

```bash
panache lint --fix document.qmd
```

After:

```markdown
# Main Title

## Skipped Level

### Another Skip
```

### Example 2: Duplicate References

Before:

```markdown
See [example1] and [example2].

[example1]: https://first.com
[example1]: https://second.com
[example2]: https://other.com
```

Run linter:

```bash
panache lint document.qmd
```

Output:

```
warning[duplicate-reference-labels]: Duplicate reference definition 'example1'
  --> document.qmd:4:1
note: First defined here:
  --> document.qmd:3:1
```

Resolution (manual):

```markdown
See [example1] and [example2].

[example1]: https://first.com
[example2]: https://other.com
```

### Example 3: Citation Validation

Before (with `refs.bib` configured):

```markdown
---
bibliography: refs.bib
---

See @existingkey and @missingkey.
```

Run linter:

```bash
panache lint document.qmd
```

Output:

```
warning[citation-keys]: Citation key 'missingkey' not found in bibliography
  --> document.qmd:5:24
```

Resolution: Add the citation to `refs.bib` or remove the reference.
